DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-12-20 2015CompleteDerivation/completeDerivation.tex
New pass at the introduction

While some branches of physics (e.g. Newtonian mechanics, thermodynamics, special relativity) are derived from physical principles (e.g. Newton's laws, laws of thermodynamics, invariance of the speed of light), others start by simply setting a mathematical framework. It is the case for classical Lagrangian mechanics, where we start by assuming the existence of a Lagrangian function whose path integral is minimized along the motion. It is the case for classical Hamiltonian mechanics, where we start by assuming the state is described by conjugate pairs $(q^i, p_i)$ and evolves according to Hamilton's equations. It is the case for quantum mechanics, where the state space is a complex inner product vector space and the state evolves according to Schroedinger's equations. That is, at the very core of some of our physical theories we have mathematical definitions instead of physical ideas.

If we ask ``when is a particular physical system suitably described by Hamiltonian (or by classical) mechanics?" the only answer we have is ``when it fits the mathematical framework". While many seem to be content with this answer (e.g. ``you have to start from somewhere, it may as well be the math which at least is precise"), it begs the question: it does not tell us how and what physical properties of the system under study are being captured.

The fundamental problem is that mathematical structures are not enough to characterize the physical system being described. For example, the same math of linear circuits can be used to describe electric, hydraulic, thermal or mechanical systems. The mathematical treatment is the same because it captures aspects of the description that are common to all cases. But for the same very reason, we cannot infer what system is being described if all we are given is the mathematical model. In the jargon of linguistic/computer science/phylosophy: math captures the syntax (the relationships between the objects) but does not capture the semantic (the meaning of each object).

We started this work with the unpretentious goal of exploring simple questions such as the following. What do these mathematical symbols used in classical/quantum Hamiltonian represent? What is the physical meaning attached to it? Why is the state of a quantum system described by a wave function (i.e. a vector in a complex inner product space), while the state of a classical hamiltonian system is described by the three pairs $(q^i, p_i)$ (i.e. a point in a cotangent bundle)? The idea was to create a math-to-physics dictionary so that each mathematical concept was clearly mapped to a crisp physical concept. But to make sure that such dictionary was complete, we realized we had to make sure that the whole relevant mathematical framework could be recovered.

@@

To our surprise, this naturally expanded the boundary of the work. The same link that allowed to justify Hamiltonian dynamics also gave us relativistic motion on the extended phase space. The same link that allowed to justify Lagrangian mechanics also onstrained the motion to that of free particles under scalar/vector potential forces. We also realized that sometimes different mathematical concepts could be justified from the same physical starting point. That is: by putting physics first we could achieve a sort of unification of ideas that was not practicable by math alone.

The success of the approach lead us to think that, maybe, the picture we were developing was not simply ``a nicer way to think about things", but instead a better framework to understand and link different parts of basic science and mathematics. After many years of work and different incremental iterations we think it is time to checkpoint and report the progress.

The goal of this paper is to show that it is possible to derive classical/quantum Hamiltonian/Lagrangian mechanics starting from a few physical principle. We do \emph{not} claim that we have perfectly achieved this result, with no room from improvement. We do claim that the whole architectural framework is solid enough to be convinced that this is indeed possible, and that it is a worthwhile endeavor as it allows to make intuitive connections between different areas of physics and mathematics. That is: it \emph{can} and \emph{should} be done.

While trying to re-derive classical, quantum and (to some extent) relativistic mechanics may seem too broad a scope, it is necessary to ensure that the concepts we develop are truly general, that they can be applied to many different cases. If we could derive classical Hamiltonian mechanics on assumptions that are inapplicable to the quantum case, what good would it do? Quantum system are Hamiltonian systems, so clearly we haven't understood what makes a system Hamiltonian. The ability to apply the same physical ideas far and wide is what gives us confidence that these are truly general concept that capture the physical aspects we are describing. The breath itself is what makes the work interesting and more than a curiosity.

As a brief overview, there are four assumptions we will consider. Assuming deterministic and reversible evolution will give us a dynamical system: a state and a law of evolution. Additionally assuming infinitesimal reducibility, studying the whole system is equivalent to studying its infinitesimal parts, gives us Hamiltonian mechanics. Additionally assuming kinematic equivalence, trajectories in space tell us everything about the state of the system, gives us Lagrangian mechanics and constrains the motion to massive particles under scalar/vector potential forces. Relativistic motion arises when studying time dependent laws of evolution with no further assumption. Alternatively, as the second assumption we can take irreducibility, studying the whole systems tells us nothing about its parts, which leads to quantum (Hamiltonian) mechanics.

@@

In section \ref{sec:dynamical_systems} we will show how the physical concept of physical distinguishability maps to the mathematical concept of topological spaces. Deterministic and reversible evolution is an invertible continuous map (i.e. a homeomorphism) as it needs to preserve what is physically distinguishable.

@@

In section \ref{sec:Lagrangian} we will introduce the assumption of kinematic equivalence, studying the motion in space is equivalent to studying the evolution of the state. Under that assumption there is a link between the state variables $(q^i, p_i)$ and the kinematic variable $(x^i, u^i)$ (i.e. position and velocity). This link will allow us to express the number of possible initial conditions for d.o.f. in terms of $dx^i$ and $du^i$ (leading to the metric tensor $g$). Such expression also allows to constrain the Hamiltonian to the one for massive particles under scalar potential forces and an equivalent Lagrangian formulation.

@@

The overall layout of the work is a balance between the need to show that the derivation follows (which requires starting from basic principles and reaching conclusion using formal arguments) and the need to give some intuitive understanding (which often requires starting from specific cases and then generalizing).

The deductive aspect is formally organized into the following:
\begin{description}
  \item[Assumptions] these characterize the physical system we are studying and constitute the premise of our discussion. A \textbf{rationale} follows each assumption, which uses physical and sometimes philosophical arguments to motivate why (or why not) such an assumption makes sense (in a particular case).
  \item[Propositions] these characterize the mathematical properties of the physical concepts we are studying. A \textbf{justification} follows each proposition to show why such characterization follows from the physical properties of the system. While the style of a justification is formal, it is not a pure mathematical proof as it necessarily has to mix physical and mathematical arguments.

@@

As these sections are clearly demarcated, this allows us to make sure that each justification only references past propositions. It also makes it easier to read the work at the desired level of detail.

@@

The work uses ideas and tools from a broad range of mathematical and scientific subjects: topology, measure theory, differential geometry (symplectic and Rimannian), group theory and vector spaces, statistics, information theory, Hamiltonian mechanics, Lagrangian mechanics, quantum mechanics, thermodynamics, special relativity and so on. Connecting these areas, in fact, is part of the insight given by this work. While this is a necessity, it does create a number of problems.

The first one is terminology and prerequisite knowledge. Since we touch a good number of areas, we cannot expect all readers to be equally comfortable with all the technical terms. Yet we cannot introduce all of them either. The compromise we adopted is to keep the informal discussion more accessible while keeping the formal derivation more precise. Even in the formal derivation, though, we will seldom need more than the basic definitions within a specific area and few major results.

The second problem is notation. As each field has developed its own set of conventions, the common usage is now an empty set. We ended up with a compromise that we believe strikes a good balance. Especially in the informal discussion, we crafted something that in general feels familiar to a physicist. Within formal parts, though, we may use notation and terminology more specific to the area of math being used. But then again: one would need to be familiar to that area anyway to be able to follow the details.

We have tested the prose on a number of students and colleagues to be confirm its accessibility at different levels. We hope the reader enjoys this journey and comes out with a few new insights and ideas.

>>>>

While some branches of physics (e.g. Newtonian mechanics, thermodynamics, special relativity) are derived from physical principles (e.g. Newton's laws, laws of thermodynamics, invariance of the speed of light), others start by simply setting a mathematical framework. It is the case for classical Lagrangian mechanics, where we start by assuming the existence of a Lagrangian function whose path integral is minimized during the motion. It is the case for classical Hamiltonian mechanics, where we start by assuming the state is described by conjugate pairs $(q^i, p_i)$ and evolves according to Hamilton's equations. It is the case for quantum mechanics, where the state space is a complex inner product vector space and the state evolves according to Schroedinger's equations. That is, at the very core of some of our physical theories we have mathematical definitions instead of physical ideas.

If we ask ``when can a particular physical system be suitably described by Hamiltonian (or by classical) mechanics?" the only answer we have is ``when it fits the mathematical framework". While many seem to be content with this answer (e.g. ``you have to start from somewhere, it may as well be the math which at least is precise"), it begs the question: it does not tell us how and what physical properties of the system under study are being captured.

This is a problem since mathematical structures are not enough to characterize the physical system being described. For example, the same math for linear circuits can be used to describe electric, hydraulic, thermal or mechanical systems. The mathematical treatment is the same because it captures aspects of the description that are common to all cases. But for the very same reason, we cannot infer what system is being described if all we are given is the mathematical model. In the jargon of linguistic/computer science/phylosophy: math captures the syntax (the relationships between the objects) but does not capture the semantic (the meaning of each object).

We started this work with the unpretentious goal of exploring simple questions such as the following. What do the mathematical symbols used in classical/quantum Hamiltonian represent? What is the physical meaning attached to it? Why is the state of a quantum particle described by a wave function (i.e. a vector in a complex inner product space), while the state of a classical particle is described by the pairs $(q^i, p_i)$ (i.e. a point in a cotangent bundle)? The idea was to create a math-to-physics dictionary so that each mathematical concept was clearly mapped to a crisp physical concept. But to make sure that such dictionary was complete, we realized we had to make sure that the whole relevant mathematical framework could be recovered.

@@

Each new link often shrank the complexity of the arguments while expanding the scope of the work in a way that was not at all anticipated. The same link that allowed to justify Hamiltonian dynamics also gave us relativistic motion on the extended phase space. The same link that allowed to justify Lagrangian mechanics also constrained the motion to that of free particles under scalar/vector potential forces. We also realized that sometimes different mathematical concepts could be justified from the same physical starting point. That is: by putting physics first we could achieve a sort of unification of ideas that was not practicable by math alone.

The unexpected success of the approach lead us to think that, maybe, the picture we were developing was not simply ``a nicer way to think about things", but instead a better framework to understand and link different parts of basic science and mathematics. After many years of work and different incremental iterations, we think the arguments are well developed enough to report the progress.

The goal of this paper is to show that it is possible to derive classical/quantum Hamiltonian/Lagrangian mechanics starting from few physical assumptions. We do \emph{not} claim that we have perfectly achieved this result, with no room from improvement. We do claim that the whole architectural framework is solid enough to be convinced that this is indeed possible, and that it is a worthwhile endeavor as it allows to make intuitive connections between different areas of physics and mathematics. That is: it \emph{can} and \emph{should} be done.

While trying to re-derive classical, quantum and (to some extent) relativistic mechanics may seem too broad a scope, it is necessary to show that the concepts we develop are truly general, that they can be applied to many different cases. If we could derive classical Hamiltonian mechanics on assumptions that are inapplicable to the quantum case, what good would it do? Quantum system are Hamiltonian systems, so clearly we wouldn't have understood what makes a system Hamiltonian. The ability to apply the same physical ideas far and wide is what gives us confidence that these are truly general concept that capture the physical aspects we are describing. The breath itself is what makes the work interesting and more than a curiosity.

As a brief overview, there are four assumptions we will consider. Assuming deterministic and reversible evolution will give us a dynamical system: states and a law of evolution. Additionally assuming infinitesimal reducibility, studying the whole system is equivalent to studying its infinitesimal parts, gives us Hamiltonian mechanics. Additionally assuming kinematic equivalence, trajectories in space tell us everything about the state of the system, gives us Lagrangian mechanics and constrains the motion to massive particles under scalar/vector potential forces. Relativistic motion arises when studying time dependent laws of evolution with no further assumption. Alternatively, as the second assumption we can take irreducibility, studying the whole systems tells us nothing about its parts, which leads to quantum (Hamiltonian) mechanics.

@@

In section \ref{sec:dynamical_systems} we will show how the physical concept of physical distinguishability leads to the mathematical concept of topological spaces. Deterministic and reversible evolution is an invertible continuous map (i.e. a homeomorphism) as it needs to preserve what is physically distinguishable.

@@

In section \ref{sec:Lagrangian} we will introduce the assumption of kinematic equivalence, studying the motion in space is equivalent to studying the evolution of the state. Under that assumption there is a link between the state variables $(q^i, p_i)$ and the kinematic variable $(x^i, u^i)$ (i.e. position and velocity). This link will allow us to express the number of possible initial conditions for a d.o.f. in terms of $dx^i$ and $du^i$, leading to the metric tensor $g$. Such expression also allows to constrain the Hamiltonian to the one for massive particles under scalar potential forces and an equivalent Lagrangian formulation.

@@

The overall layout of this work is a balance between the need to show that the derivation follows (which requires starting from basic principles and proceeding using formal arguments) and the need to give an intuitive understanding (which often requires starting from specific cases and then generalizing).

The deductive aspect is formally organized into the following:
\begin{description}
  \item[Assumptions] these characterize the physical system we are studying and constitute the premise of our discussion. A \textbf{rationale} follows each assumption, which uses physical and sometimes philosophical arguments to motivate why (or why not) such an assumption makes sense (in a particular case).

  \item[Propositions] these characterize the mathematical properties of the physical concepts we are studying. A \textbf{justification} follows each proposition to show why such characterization captures the particular physical properties of the system. While the style of a justification is formal, it is not a pure mathematical proof as it necessarily has to mix physical and mathematical arguments.

@@

As these sections are clearly demarcated, they allow us to make sure that each justification only references past propositions. It also makes it easier to read the work at the desired level of detail.

@@

The work uses ideas and tools from a broad range of mathematical and scientific subjects: topology, measure theory, differential geometry (symplectic and Rimannian), group theory and vector spaces, statistics, information theory, Hamiltonian mechanics, Lagrangian mechanics, quantum mechanics, thermodynamics, special relativity and so on. Using idea from all these areas is necessary, in fact unraveling their connections is part of the value of this work, but it does create a number of problems.

The first one is terminology and prerequisite knowledge. Since we touch a good number of areas, we cannot expect all readers to be equally comfortable with all the technical terms. Yet we cannot introduce all of them either. The compromise we adopted is to keep the informal discussion more accessible while keeping the formal derivation more precise, without needing more than the basic definitions and few major results within each specific area. This may leave an expert in a particular field not fully satisfied but, as we mentioned in the introduction, our goal is to lay out the general picture.

The second problem is notation. As each field has developed its own set of conventions, the common usage is now an empty set. We ended up with a compromise that hopefully feels natural enough to a broad audience of physicist, mathematical physicists and philosophers of physics. Within formal parts we may use notation and terminology more specific to the area of math being used, but one would need to be familiar with them anyway to be able to follow the details.

We have tested the prose on a number of colleagues and students to confirm its accessibility at different levels and believe we have struck a decent compromise. We hope the reader enjoys this journey and comes out with a few new insights and ideas.
